---
title: An Introduction to Shiny
author: Joel Chaney
date: Cin-Day RUG / 25SEP2019
framework: html5slides
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

## What is Shiny and How Can It Help Me?
- Shiny is an R package developed, maintained and supported by RStudio 
    - <http://www.rstudio.com/shiny/>
        - General information
        - Tutorials
        - Server implementation options
- Shiny allows you to create interactive statistical web tools ***easily*** and with great ***flexibility*** and ***power***
    - Incorporate analyses and graphs from one or more of nearly 15,000 R packages now available
    - Create interactive graphics and tables using only R code - no additional scripting necessary!
        - **ui.R**:  An R script to create the user interface
        - **server.R**:  An R script to conduct analyses and create graphs and/or tables for display
        - **app.R**:  A single R script used instead of ui.R and server.R

## How Do I Obtain "Shiny"?
- Shiny is available on CRAN (the Comprehensive R Archive Network), so simply do one of the following:
    - In RStudio, click on "Tools" > "Install Packages...", type "shiny" under "Packages" and click the "Install" button
    - Click on "Packages" > "Install Packages..." in the R command window, select a CRAN mirror and then select "shiny"
    - Type "install.packages('shiny')" in the R command window

## How Do I Start Using "Shiny"?
- Go through the video and/or written tutorials at <http://shiny.rstudio.com/tutorial/> 
- Set up the tool:
    - manually via the R command window or GUI, or
    - via RStudio
    
## How Do I Start Using "Shiny" (with RStudio)?
- Create a new Shiny app:
    - Select "File" > "New Project..." > "New Directory" > "Shiny Web Application"
    - Enter a new folder name under "Directory name:"
    - Browse to a location to place the folder
    - Click "Create Project"
- ui.R and server.R files are created in the new folder
- Click the "Run App" button at the top of the script window to run the example tool created
- Modify and save the ui.R and server.R scripts to create a new tool

## How Do I Start Using "Shiny" (without RStudio)?
- Select "Packages" > "Load Package." > "shiny", or type "library(shiny)", in the R command window after the package has been installed
- Place an "app.R" file, or both a "ui.R" file and a "server.R" file, in your working directory 
    - Type "getwd()" to see what your working directory is
    - Type "setwd('\<location\>')" to change your working directory to <location> (e.g., 'U:/Data/MyShinyApps')
- Type "runApp()" on the command line - the tool will start in your default web browser

## User Interface R File (ui.R)
- Use the shinyUI() function, followed by the fluidPage() function
- Within fluidPage():
    - titlePanel()
    - sidebarLayout()
        - sidebarPanel()
        - mainPanel()
- HTML builder functions:
    - HTML(), p(), br(), h1(), ..., h6(), a(), ...
- Input control/widget functions:
    - numericInput(), sliderInput(), fileInput(), ...
- Output functions:
    - downloadButton(), plotOutput(), tableOutput(), dataTableOutput(), ...

## Server R File (server.R)
- Include the analysis/graph code in the shinyServer() function
- Within shinyServer():
    - reactive():  used to automatically update object values
    - renderPlot():  display a reactive plot
    - renderImage():  display an image
    - renderTable():  display a data table
    - renderDataTable():  display an interactive data table
    - downloadHandler():  make file download available

## Example 1: Getting Our Feet Wet with Geyser Data
- Begin with the RStudio starting app
- Modify the slider so that values from 5 to 25 can be selected for the # of bins
- Change the starting # of bins value to 15
- Make the width of the sidebar panel to 3 units rather than the default of 4 units
- Change the bar color to your favorite color

## Example 1: ui.R
```{r, echo=TRUE, eval=FALSE}
library(shiny)
shinyUI(fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),
  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
                  "Number of bins:",
                  min = 5,
                  max = 25,
                  value = 15), 
      width = 3
    ),
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("distPlot"),
      width = 9
    )
  )
))
```  

## Example 1: server.R
```{r, echo=TRUE, eval=FALSE}
library(shiny)
shinyServer(function(input, output) {
  output$distPlot <- renderPlot({
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    # draw the histogram with the specified number of bins
    hist(x, breaks = bins, col = 'blueviolet', border = 'white')
  })
})
```  

## Example 2: Exact Binomial Test
- Numeric and radio button input in addition to slider input
- Renders a table and a static image
- Includes a download button (to get analysis results in a .csv file)
- Includes a reactive() function 

## Example 2: ui.R
```{r, echo=TRUE, eval=FALSE}
library(shiny)
shinyUI(fluidPage(
  titlePanel("Test for One Proportion (Exact Binomial Test)"),
  sidebarLayout(
    # Sidebar with controls to select the # of successes, sample size, confidence level and alternative hypothesis
    sidebarPanel(
      p("Enter the number of \"successes\" and the total sample size for a single group below, and enter the desired confidence level and alternative hypothesis 
        (i.e., statement you would like to make).  The table to the right will be automatically updated with the statistical test results.  This table can be downloaded
        in .csv format by clicking on the \"Download the Results\" button."),
      fluidRow(
        column(6, numericInput("s1", "# of Successes:", 12)),
        column(6, numericInput("t1", "Total Sample Size:", 30))
      ),
      sliderInput("confid", "Confidence Level (%):", 50, 99, 95, 1), 
      radioButtons("testType", "Alternative Hypothesis:", c("proportion of successes is not equal to 0.5"="two.sided", "proportion of successes is less than 0.5"="less", 
        "proportion of successes is greater than 0.5"="greater"))
    ),
    # Show the results table and download button
    mainPanel(
      h3(textOutput("Exact Binomial Test")),
      tableOutput("stats"),
      downloadButton('downloadResults', 'Download the Results')
    )
  )
))
```  

## Example 2: server.R
```{r, echo=TRUE, eval=FALSE}
library(shiny)
shinyServer(function(input, output) {
  altHypothesisLabel <- reactive({ ifelse(input$testType=="two.sided", "Not Equal", ifelse(input$testType=="less", "Less Than", "Greater Than")) })
  propTest <- function() {
    results <- binom.test(input$s1, input$t1, conf.level=input$confid/100, alternative=input$testType)
    s1 <- sprintf("%1.0f", input$s1)
    t1 <- sprintf("%1.0f", input$t1)
    s2 <- sprintf("%1.0f", input$t1 - input$s1)
    prop1 <- sprintf("%1.3f", input$s1/input$t1)
    prop2 <- sprintf("%1.3f", (input$t1 - input$s1)/input$t1)
    propdiff <- sprintf("%1.3f", input$s1/input$t1 - (input$t1 - input$s1)/input$t1)
    pvalue <- sprintf("%1.4f", results$p.value)
    c1 <- sprintf("%1.3f", results$conf.int[1])
    c2 <- sprintf("%1.3f", results$conf.int[2])
    clevel <- paste0(sprintf("%1.0f", input$confid), "%")
    out <- data.frame(paste0(prop1, "  (", s1, "/", t1, ")"), paste0(prop2, "  (", s2, "/", t1, ")"), pvalue, c1, c2, altHypothesisLabel(), clevel)
    names(out) <- c("Success Proportion", "Failure Proportion", "p", "Success Proportion Lower CL", "Success Proportion Upper CL", "Alternative Hypothesis", "Confidence Level")
    return(out)
  }
  output$stats <- renderTable({ propTest() }, align=rep("c", 8), include.rownames=FALSE)
  output$downloadResults <- downloadHandler(
    filename = "output.csv",
    content = function(file) {
      write.csv(propTest(), file, row.names=FALSE)
    }
  )
})
```  

## Example 3: Exact Binomial Test with a Graph
- Let's add a graph to the exact binomial test app
- Graph shows:
    - the binomial distribution for a given total sample size
    - highlighted bar representing the # of successes
- In ui.R, use "plotOutput()" (analogous to tableOutput())
- In server.R, use "renderPlot()" (analogous to renderTable())
- Use a function we create called graphBinom()

## Example 3: Code added to mainPanel in ui.R
```{r, echo=TRUE, eval=FALSE}
		mainPanel(
			h3(textOutput("Exact Binomial Test")),
			tableOutput("stats"),
			br(),
			plotOutput("graph"),
			downloadButton('downloadResults', 'Download the Results')
		)
```  

## Example 3: Code added to server.R
```{r, echo=TRUE, eval=FALSE}
  graphBinom <- function(x, totaln) {
    pbin <- pbinom(1:totaln, size=totaln, prob=0.5)
    prev <- c(0, pbin[-length(pbin)])
    indiv <- pbin-prev
    cols <- rep("blue", totaln)
    cols[x] <- "red"
    xval <- barplot(indiv, col=cols, xlab="# of Successes", ylab="Probability")
    axis(1, xval, labels=1:totaln)
  }  

  output$graph <- renderPlot({ graphBinom(input$s1, input$t1) })
```  

## Example 4: Reactive Functions
- Purpose: Prevent functions from being run repeatedly/unnecessarily
- Replace 
```{r, echo=TRUE, eval=FALSE}
propTest <- function() { ... }
```  
with
```{r, echo=TRUE, eval=FALSE}
propTest <- reactive({ ... })
```  
- In this example, the file download occurs immediately, without the need to re-run propTest()
- Alternatively:
```{r, echo=TRUE, eval=FALSE}
propTest <- function() { ... }
propTestR <- reactive({ propTest() })
```  
then call propTestR() instead of propTest()

## Example 5: Interactive Graphs
- So far, we've seen interactivity via widgets
- Graphs that are truly interactive can be created in Shiny by integrating other packages, such as:
    - Plotly
    - Bokeh
    - ggplot2
- Let's look at a Plotly example
- Install the "plotly" package:
    - In RStudio: Packages > Install > plotly, or type
```{r, echo=TRUE, eval=FALSE}
install.packages("plotly")
```  
in the console window

## Example 5: Interactive Graphs
Replace the bar chart with an interactive Plotly bar chart
In ui.R, replace 
```{r, echo=TRUE, eval=FALSE}
			plotOutput("graph"),
```  
with
```{r, echo=TRUE, eval=FALSE}
			plotlyOutput("graph"),
```  
In server.R, replace 
```{r, echo=TRUE, eval=FALSE}
  output$graph <- renderPlot({ graphBinom(input$s1, input$t1) })
```  
with
```{r, echo=TRUE, eval=FALSE}
  output$graph <- renderPlotly({ graphBinom(input$s1, input$t1) })
```  

## Example 5: Interactive Graphs
In the graphBinom() function, replace:
```{r, echo=TRUE, eval=FALSE}
    xval <- barplot(indiv, col=cols, xlab="# of Successes", ylab="Probability")
    axis(1, xval, labels=1:totaln)
```  
with 
```{r, echo=TRUE, eval=FALSE}
    plot_ly(x=1:totaln, y=indiv, type="bar", marker=list(color=toRGB(cols))) %>% 
      layout(xaxis=list(title="# of Successes"), 
             yaxis=list(title="Probability"))
```  

## Example 6: Uploading Files / Displaying Interactive Data Tables
- Files can be easily uploaded in Shiny via fileInput() in ui.R
- Data can be displayed in a static table (as we've already seen) or within interactive data tables
- Interactive data tables are based on the JavaScript DataTables library (<https://datatables.net/>)
- dataTableOutput() instead of tableOutput() in ui.R
- renderDataTable() instead of dataTable() in server.R

## Example 6: Uploading Files / Displaying Interactive Data Tables
ui.R
```{r, echo=TRUE, eval=FALSE}
library(shiny)
shinyUI(fluidPage(
  titlePanel("Data Viewer"),
  sidebarLayout(
    sidebarPanel(
      fileInput("file1", "Choose a .csv file", multiple=FALSE, accept=".csv"),
      tags$hr(),
      checkboxInput("header", "Check if the first row is a header", TRUE)
    ),
    mainPanel(
      dataTableOutput("contents")
    )
  )
))
```  

## Example 6: Uploading Files / Displaying Interactive Data Tables
server.R
```{r, echo=TRUE, eval=FALSE}
library(shiny)
shinyServer(function(input, output) {
  output$contents <- renderDataTable({
    inFile <- input$file1
    if (is.null(inFile))  return(NULL)
    if (!is.null(inFile)) {
      indata <- read.csv(inFile$datapath, header=input$header)
  	  return(indata)
    }  
  })
})
```  

## Example 7: Displaying Progress Information
- Progress bars and messages can be displayed in apps
- Useful to let end-users know that something is happening
- withProgress() and setProgress() functions

## Example 7: Displaying Progress Information
server.R
```{r, echo=TRUE, eval=FALSE}
shinyServer(function(input, output, session) {
  altHypothesisLabel <- reactive({ ifelse(input$testType=="two.sided", "Not Equal", ifelse(input$testType=="less", "Less Than", "Greater Than")) })
  propTest <- function() {
    # wait for no good reason other than to demonstrate the "reactive" function
    withProgress(session, min=0, max=100, {
      setProgress(message="Sleeping for 10 seconds...", detail="Please wait.", value=25)
      Sys.sleep(10)
      # now do the analysis
      setProgress(message="Conducting analysis...", value=80)
      ...
  	  setProgress(message="Done.", value=95)
    })  # end withProgress
    return(out)
  }
```  

## Shiny Tool Deployment
- "Shiny Server" software can be deployed on Linux servers
    - open-source version is available on RStudio's website
    - RStudio licenses "Shiny Server Pro" specifically for Shiny app deployment
    - RStudio licenses "RStudio Connect" for Shiny app deployment and other R-based content
- Without Shiny Server:
    - share the ui.R and server.R files, or app.R, with others who have R and Shiny installed
    - default web browser and localhost are used
